# Analysis

This chapter will clean the data further and perform interpolation on each well from the chosen set. First, the Extreme Learning Machine method, described in Chapter 3, will be used to impute missing values. A spatial interpolation method, Inverse Distance Weighting (IDW) will then be used for comparison purposes. Finally, the accuracy of each method will be analyzed and compared.

In order to determine accuracy, three years (2012-2015) of data will be imputed by each method. Then, the imputed data will be compared to the observed data.

## Extreme Learning Machine

As described previously, the ELM method forms a relationship between input data sources (in our case, remotely sensed earth observation data) and observed data. This relationship is then used to impute data for time periods without measurements.

First, we will need to create a time series data frame that contains all the measurements from each well. We will summarize the data by year to simplify things.
```{r message=FALSE}
load("code/master.Rda")
master_wells <- read.csv(file = 'code/master_wells.csv')

test_wells <- c('25N03W11B003M', 
                '29N03W18M001M', 
                '24N02W29N004M', 
                '24N02W24D003M',
                '24N02W03B001M')
test_wells <- subset(master_wells, mergeOn %in% test_wells)

test_ts_yr <- data.frame(c())
for (well in test_wells$mergeOn){
  ts <- filter(master, mergeOn == well)
  yr_ts <- ts %>%
    mutate(date = lubridate::floor_date(date, "1 year"), mergeOn = mergeOn) %>%
    group_by(date, mergeOn) %>%
    summarize(Mean_depth=-1*mean(depth.to.GW..ft.))
  test_ts_yr <- as.data.frame(rbind(test_ts_yr, yr_ts))
  #assign(paste(well, "_3mon_ts", sep = ""), mon_ts)
}
```

We can also create a facet-wrapped group of time series graphs for each well.
```{r}
# Now let's create time series graphs of these yearly means
test_ts_yr %>%
  ggplot(aes(x = date, y = Mean_depth)) + 
  geom_line() +
  xlab("Date") +
  ylab("Depth (ft)") +
  ggtitle(well) +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~mergeOn)
```

## Example two

## Example three
